### 2022-03-29 manuscript.tex
Update 02.tubaseq.md [ci skip]
This build is based on
https://github.com/scherzerthesis/thesis/commit/02d58ac3e5f0bd59f708837074746dc2eb309ec4.

This commit was created by the following CI build and job:
https://github.com/scherzerthesis/thesis/commit/02d58ac3e5f0bd59f708837074746dc2eb309ec4/checks
https://github.com/scherzerthesis/thesis/actions/runs/2055574609

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/a6a894e99cfd62a104f67d3ecf75754b618afe0c/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/a6a894e99cfd62a104f67d3ecf75754b618afe0c}{scherzerthesis/thesis@a6a894e}
on March 28, 2022.

@@

Indeed, in highly mutated cancers such as melanoma and lung cancer, recurrent mutations are not necessarily predictive of biological importance. To better understand the importance of candidate tumor suppressors or oncogenes, researchers exploit tumor cell lines and GEMMs to identify bona-fide genetic drivers of cancer. However, these systems are suboptimal, as they are either not physiologically relevant (tissue culture) or are not scalable to determine the effect of many mutations (GEMMs). Moreover, techniques to measure cooperation of genes to drive lung cancer are qualitative and are not sufficiently quantitative to allow for large-scale studies. typically, lung adenocarcinoma is quantified by lung weight or histological analyses such as tumor burden (Percentage of lung that is tumor), tumor size, and immunohistochemistry (IHC) of tumor markers. Although important and informative, these analyses are not high-throughput and rely on many mice to acheive statistical confidence. Furthermore, there is intrinsic bias when calculating tumor size and tumor burden, as tissue-sectioning results in tumor sections that do not accurately portray true tumor size. Therefore, more advanced techniques are required for large-scale analysis of mouse tumors that are both quantitative and high throughput.

@@

BRAF\textsuperscript{V600E} has previously been shown to be a potent driver of lung tumorigenesis(Dankort \emph{et al.}, 2007). However, tumor fails to progress past adenoma-like tumors. Given the theory that BRAF-\textsuperscript{V600E} induces growth arrest after an initial burst of neoplastic growth, it would be of interest to know if all tumors grow to a predetermined and absolute size, or if there is a spectrum of tumor sizes that are determined by extrinsic factors such as proximity to blood vessels or certain stromal cells.Therfore, we used Tuba-Seq combined with CRISPR/CAS9 genome editing to explore the contribution of 11 tumor suppressor genes and 5 controls in the context of our BRAF \textsuperscript{V600E}- driven lung tumor models.

@@

Given our findings of SETD2 loss increased tumor burden concurrent with BRAF\textsuperscript{V600E} we next validated SETD2 in a new cohort of mice with single-guide RNAs targeting SETD2. Indeed, loss of SETD2 increased tumor burden from 5\% to over 30\% in some animals. Since SETD2 antibodies are unreliable in both western blots and immunohistochemistry we used H3K36me3 as a surrogate marker for SETD2 activity by immunohistochemistry. Whereas 90\% of nuclei in BRAF \textsuperscript{V600E} only tumors were H3K36me3, more than 60\% tumors of mice infected with SETD2 CRISPR virus were negative for H3K36me3. Furthermore, when classified by H3K36me3-positive or -negative, there is even stronger evidence that SETD2 loss leads to larger tumor formation

@@

The mutational landscape of lung adenocarcinoma is complex due lifelong bombardment on DNA by carcinogen ingested into the distal lung epithelium{[}23770567{]}. While understanding which altered genes play causal roles in tumor formation and progression is required for personalized medicine, the noisy mutation profiles in lung adenocarcinoma makes it hard to do so. However, it is clear that there is a central role for the MAPK pathway during the initiation and progression of lung adenocarcinoma (Cicchini \emph{et al.}, 2017). Oncogenic BRAF\textsuperscript{V600E} is a major driver of lung adenocarcinoma. However, mouse models with activation of BRAF \textsuperscript{V600E} only lead to benign tumorigenesis(Shai \emph{et al.}, 2015; Dankort \emph{et al.}, 2007). Although it is known that activation of WNT and PI3k signaling, as well as lost of \emph{Trp53} and \emph{Cdkn2a} genes allow BRAF\textsuperscript{V600E} benign lung tumors to progress to malignant disease, it is a time-consuming and laborious process to effects of individual tummor suppressors in GEMMs.Therefore, we sought out to systematically test a variety of known tumor suppressors in lung adenocarcinoma. Furthermore,untilnow, we did not know the absolute number of cells in a BRAF\textsuperscript{V600E} adenoma, only its area, which is confounded by tissue sectioning as well as stromal cell contamination. Prior to Tuba-Seq, it was very difficult to accurately determine the size of BRAF\textsuperscript{V600E} adenomas. Based on population doublings, as well as cross sections of BRAF \textsuperscript{V600E} tumors in histological sections, it is estimated that BRAF\textsuperscript{V600E}-driven lung adenomas grow about 15-20 populations doublings and reaching a size of 1mm\textsuperscript{3}(Juan \emph{et al.}, 2014). With Tuba-Seq, we accurately discerned the absolute size of all BRAF\textsuperscript{V600E} lung adenomas, regardless of size. Interestingly, our results have revealed BRAF\textsuperscript{V600E}-driven tumors widely differ in absloute tumor cell number. Our results support the hypothesis that each tumor-initiating cell has its own unique barriers for tumor outgrowth. This heterogenous reponse to BRAF\textsuperscript{V600E} activation likely reflects intrinisic differences between individual cell states. Indeed, tumorigenic potential can vary greatly due to different cell types (AT2 vs Club cell vs AT1), or can be influenced by subtle differences within a cell type, such proximity to supporting stromal cells such as fibroblast or immune cells. Since we used a ubiquitous promoter to drive BRAF\textsuperscript{V600E} and CAS9 expression, it is likely that not all sequence reads come from adenomas or tumors, but rather from normal cell turnover during tissue homeostasis. Therefore, in out quantifcations and analysis, we focused on tumors that were greater than 5,000 cells.

@@

We have also identified \emph{SETD2} loss as a potent tumor supressor(Figure 2). \emph{SETD2} is mutated in other solid malignancies on cBioPortal, such as Kidney and bladder cancers(Cerami \emph{et al.}, 2012). SETD2 is a histone methyltransferase that deposits trimethylation marks on H3K36(Bhattacharya and Workman, 2020). H3K36me\textsuperscript{3} is used by chromatin modifiers to mark actively trascribed genes(Edmunds \emph{et al.}, 2007). Data mining of TCGA mRNA expression and survival analysis further revealed that \emph{SETD2} mRNA is lower in lung adenocarcinoma and is associated with poorer overall survival(Figure 2). Interestingly, \emph{SETD2} alterations are also significantly associated with BRAF\textsuperscript{V600E} mutations (Figure 2). The unique association of \emph{BRAF\textsuperscript{V600E}} mutation and \emph{SETD2} loss may reflect an interesting synergistic effect between MAPK signaling and loss of chromatin regulation. One hypothesis is that SETD2 loss drives spurious transcriptional activation. Similarily, Recent work by Monte Winslow and also David Feldser's labs have indentified SETD2 loss as a cooperating event in KRAS-G12D lung tumorigenesis(Walter \emph{et al.}, 2017). However, the mechanism by which \emph{SETD2} loss drives tumor progression in either BRAF\textsuperscript{V600E} or KRAS-G12D remain elusive.

>>>>

This manuscript
(\href{https://scherzerthesis.github.io/thesis/v/02d58ac3e5f0bd59f708837074746dc2eb309ec4/}{permalink})
was automatically generated
from \href{https://github.com/scherzerthesis/thesis/tree/02d58ac3e5f0bd59f708837074746dc2eb309ec4}{scherzerthesis/thesis@02d58ac}
on March 29, 2022.

@@

Indeed, in highly mutated cancers such as melanoma and lung cancer, recurrent mutations are not necessarily predictive of biological importance.
To better understand the importance of candidate tumor suppressors or oncogenes, researchers exploit tumor cell lines and GEMMs to identify bona-fide genetic drivers of cancer.
However, these systems are suboptimal, as they are either not physiologically relevant (tissue culture) or are not scalable to determine the effect of many mutations (GEMMs).
Moreover, techniques to measure cooperation of genes to drive lung cancer are qualitative and are not sufficiently quantitative to allow for large-scale studies.
Typically, lung adenocarcinoma is quantified by lung weight or histological analyses such as tumor burden (Percentage of lung that is tumor), tumor size, and immunohistochemistry (IHC) of tumor markers.
Although important and informative, these analyses are not high-throughput and rely on many mice to acheive statistical confidence. Furthermore, there is intrinsic bias when calculating tumor size and tumor burden, as tissue-sectioning results in tumor sections that do not accurately portray true tumor size. Therefore, more advanced techniques are required for large-scale analysis of mouse tumors that are both quantitative and high throughput.

@@

BRAF\textsuperscript{V600E} has previously been shown to be a potent driver of lung tumorigenesis(Dankort \emph{et al.}, 2007).
However, tumor fails to progress past adenoma-like tumors.
Given the theory that BRAF-\textsuperscript{V600E} induces growth arrest after an initial burst of neoplastic growth, it would be of interest to know if all tumors grow to a predetermined and absolute size, or if there is a spectrum of tumor sizes that are determined by extrinsic factors such as proximity to blood vessels or certain stromal cells.
Therfore, we used Tuba-Seq combined with CRISPR/CAS9 genome editing to explore the contribution of 11 tumor suppressor genes and 5 controls in the context of our BRAF \textsuperscript{V600E}- driven lung tumor models.

@@

Given our findings of SETD2 loss increased tumor burden concurrent with BRAF\textsuperscript{V600E} we next validated SETD2 in a new cohort of mice with single-guide RNAs targeting SETD2.
Indeed, loss of SETD2 increased tumor burden from 5\% to over 30\% in some animals.
Since SETD2 antibodies are unreliable in both western blots and immunohistochemistry we used H3K36me3 as a surrogate marker for SETD2 activity by immunohistochemistry.
Whereas 90\% of nuclei in BRAF \textsuperscript{V600E} only tumors were H3K36me3, more than 60\% tumors of mice infected with SETD2 CRISPR virus were negative for H3K36me3.
Furthermore, when classified by H3K36me3-positive or -negative, there is even stronger evidence that SETD2 loss leads to larger tumor formation.

@@

The mutational landscape of lung adenocarcinoma is complex due lifelong bombardment on DNA by carcinogen ingested into the distal lung epithelium{[}23770567{]}.
While understanding which altered genes play causal roles in tumor formation and progression is required for personalized medicine, the noisy mutation profiles in lung adenocarcinoma makes it hard to do so.
However, it is clear that there is a central role for the MAPK pathway during the initiation and progression of lung adenocarcinoma (Cicchini \emph{et al.}, 2017).
Oncogenic BRAF\textsuperscript{V600E} is a major driver of lung adenocarcinoma.
However, mouse models with activation of BRAF \textsuperscript{V600E} only lead to benign tumorigenesis(Shai \emph{et al.}, 2015; Dankort \emph{et al.}, 2007).
Although it is known that activation of WNT and PI3k signaling, as well as lost of \emph{Trp53} and \emph{Cdkn2a} genes allow BRAF\textsuperscript{V600E} benign lung tumors to progress to malignant disease, it is a time-consuming and laborious process to effects of individual tummor suppressors in GEMMs.
Therefore, we sought out to systematically test a variety of known tumor suppressors in lung adenocarcinoma.
Furthermore,until now, we did not know the absolute number of cells in a BRAF\textsuperscript{V600E} adenoma, only its area, which is confounded by tissue sectioning as well as stromal cell contamination.
Prior to Tuba-Seq, it was very difficult to accurately determine the size of BRAF\textsuperscript{V600E} adenomas.
Based on population doublings, as well as cross sections of BRAF \textsuperscript{V600E} tumors in histological sections, it is estimated that BRAF\textsuperscript{V600E}-driven lung adenomas grow about 15-20 populations doublings and reaching a size of 1mm\textsuperscript{3}(Juan \emph{et al.}, 2014).
With Tuba-Seq, we accurately discerned the absolute size of all BRAF\textsuperscript{V600E} lung adenomas, regardless of size.
Interestingly, our results have revealed BRAF\textsuperscript{V600E}-driven tumors widely differ in absloute tumor cell number. Our results support the hypothesis that each tumor-initiating cell has its own unique barriers for tumor outgrowth.
This heterogenous reponse to BRAF\textsuperscript{V600E} activation likely reflects intrinisic differences between individual cell states.
Indeed, tumorigenic potential can vary greatly due to different cell types (AT2 vs Club cell vs AT1), or can be influenced by subtle differences within a cell type, such proximity to supporting stromal cells such as fibroblast or immune cells.
Since we used a ubiquitous promoter to drive BRAF\textsuperscript{V600E} and CAS9 expression, it is likely that not all sequence reads come from adenomas or tumors, but rather from normal cell turnover during tissue homeostasis.
Therefore, in out quantifcations and analysis, we focused on tumors that were greater than 5,000 cells.

@@

We have also identified \emph{SETD2} loss as a potent tumor supressor(Figure 2).
\emph{SETD2} is mutated in other solid malignancies on cBioPortal, such as Kidney and bladder cancers(Cerami \emph{et al.}, 2012).
SETD2 is a histone methyltransferase that deposits trimethylation marks on H3K36(Bhattacharya and Workman, 2020).
H3K36me\textsuperscript{3} is used by chromatin modifiers to mark actively trascribed genes(Edmunds \emph{et al.}, 2007).
mining of TCGA mRNA expression and survival analysis further revealed that \emph{SETD2} mRNA is lower in lung adenocarcinoma and is associated with poorer overall survival(Figure 2).
Interestingly, \emph{SETD2} alterations are also significantly associated with BRAF\textsuperscript{V600E} mutations (Figure 2).
The unique association of \emph{BRAF\textsuperscript{V600E}} mutation and \emph{SETD2} loss may reflect an interesting synergistic effect between MAPK signaling and loss of chromatin regulation.
One hypothesis is that SETD2 loss drives spurious transcriptional activation.\\
Similarily, Recent work by Monte Winslow and also David Feldser's labs have indentified SETD2 loss as a cooperating event in KRAS-G12D lung tumorigenesis(Walter \emph{et al.}, 2017).
However, the mechanism by which \emph{SETD2} loss drives tumor progression in either BRAF\textsuperscript{V600E} or KRAS-G12D remain elusive.
